Students' Alcohol Consumption Impacts on Academic Results on the Example of a Secondary Scool Maths Class in Portugal

### 2016-10-14 knit_MSproject.Rmd
small change
Add files via upload
edited regression part + added correlation part as intro to regression (only high correlations between y and x's)

    + a. General Information
    + b. Family
    + c. Education 
    + d. Social Life and Health
    + e. Grades  \n

@@

## 1. Introduction 
Recent events in the world economy have lead to the decline in economic performance in many countries, amongst which is the decline of econommic growth in the European Union (EU).  As a response to these changes,  the EU has made it a priority to reverse this trend through investments that would provide new jobs and create a better educated workforce. Hence, the European Commission (EC) has set up a ‘Strategy for smart, sustainable and inclusive growth’ (European Comission, 2010), where a crucial part of investment is to contribute to the education of the next generations: The aim is to reduce the share of early school leavers to below 10% and ensure that at least 40% of the younger generation obtain a tertiary degree.

@@

## 2. Dataset Description and Descriptive Statistics 
```{r a, include=FALSE}
# Read csv file
sm <- read.csv("/Users/Felix/Desktop/R_project/student-mat.csv", sep=';')

@@

#### d. Social Life and Health 
The social life section provieds information on activities the student is involved in outside of school. This includes how much leisure time they have, whether they are in a relationship and how often they go out. Moreover, there is data on how much alcohol students consume and in what health condition they are in.

#### e. Grades
The grades are split into three grades: the first period (G1), second period (G2) and final grade (G3). For the purpose of thy analysis, G1 and G2 will be neglected. In addition, there is information on whether they have failed to pass any classes. 
```{r e, echo=FALSE}
pie_failures
```

@@

@@

## 4. Correlation Analysis and Regression
In this report, we would like to find factors that are related to students academic performance. We will perform a multivariate regression analysis and use step-wise variable selection to find significant variables related to our response variable (academic performance - "Score"). Prior to the regression analysis, we want to get a general idea about the relationship between the variables. There are several things we want to inspect:

a. Distribution of response variable

if there is skewness in the distribution, we should consider transforming the variable (i.e logarithmic)

b. Correlation between response variable and each explanatory variables

c. Correlation among explanatory variables

if there is a high correlation among explanatory variables, then there is possible multicollinearity problem

After performing the inspection and do the necessary action, we will continue with the regression and check the validity of the model.

### a. Distribution of Response Variable
```{r f, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)

sm[,"Score"] <- (sm[,"G1"] + sm[,"G2"] + sm[,"G3"])/3

ggplot(sm) + geom_density(aes(x=Score)) + labs(x = "Academic Score", y="% Frequency", title = "Academic Score of Secondary School Student")
```
From the graph, we can see that the distribution is slightly left-skewed. Generally, we would transform the variable to become more normally distributed. However, for the purpose of this project, we will not do any transfomation on the response variable and assume that the distribution is normal.

### b. Correlation between response variable and each explanatory variables
```{r g, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}

#below - too large to knit

#pairs(sm[,c("G3", "G1","G2","school","sex","age", "address","famsize","Pstatus","Medu","Fedu",
#            "Mjob","Fjob","reason","guardian","traveltime","studytime","failures",
#            "schoolsup","famsup","paid","activities","nursery","higher","internet",
#            "romantic","famrel","freetime","goout","Dalc","Walc","health","absences")],
#      lower.panel = panel.smooth, upper.panel = panel.cor)

@@

### c. Correlation between response variable and each explanatory variables
```{r h, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}

# too large to knit

#pairs(sm[,c("G3", "G1","G2","school","sex","age", "address","famsize","Pstatus","Medu","Fedu",
#            "Mjob","Fjob","reason","guardian","traveltime","studytime","failures",
#            "schoolsup","famsup","paid","activities","nursery","higher","internet",
#            "romantic","famrel","freetime","goout","Dalc","Walc","health","absences")],
#      lower.panel = panel.smooth, upper.panel = panel.cor)
```

```{r i, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#2. derive variables

# father or mother job is teacher = 1
sm$f_or_mteach <- 0
sm[j,"f_or_mteach"] <- as.character(sm[,"f_or_mteach"])
for (j in 1:nrow(sm)) {

  if (sm[j, "Fjob"] == "teacher") sm[j,"f_or_mteach"] <- 1
  if (sm[j, "Mjob"] == "teacher") sm[j,"f_or_mteach"] <- 1
}

sm[,"f_or_mteach"]

pairs(sm[,c("G3","f_or_mteach")], lower.panel = panel.smooth, upper.panel = panel.cor)

# father or mother earns university degree = 1 
sm$f_or_medu4 <- 0
sm[j,"f_or_medu4"] <- as.character(sm[,"f_or_medu4"])
for (j in 1:nrow(sm)) {

  if (sm[j, "Medu"] == 4) sm[j,"f_or_medu4"] <- 1
  if (sm[j, "Fedu"] == 4) sm[j,"f_or_medu4"] <- 1
}

sm[,"f_or_medu4"] 
pairs(sm[,c("G3","f_or_medu4")], lower.panel = panel.smooth, upper.panel = panel.cor)

### Regression
```{r j, include=FALSE}

# academic performance is measured by the average of exam scores
sm[, "Score"] <- (sm[, "G1"] + sm[, "G2"] + sm[, "G3"])/3

#Initial regression

reg <- lm(Score ~ school + sex + age + address + famsize + Pstatus + Medu + Fedu + 
                Mjob + Fjob + reason + guardian + traveltime + studytime + failures + 
                schoolsup + famsup + paid + activities + nursery + higher + internet + 
                romantic + famrel + freetime + goout + Dalc + Walc + health + absences +
                f_or_mteach + f_or_medu4
                , data = sm)
summary(reg)

@@

#To cross check on the final model and the significant variables, we use existing R library:

library(MASS)
step <- stepAIC(reg, direction="both")

#Final Model

@@

>>>>

    + a. General Information
    + b. Family
    + c. Education 
    + d. Social Life and Health
    + e. Grades

@@

## 1. Introduction 
Recent events in the world economy have lead to the decline in economic performance in many countries, amongst which is the decline of economic growth in the European Union (EU).  As a response to these changes,  the EU has made it a priority to reverse this trend through investments that would provide new jobs and create a better educated workforce. Hence, the European Commission (EC) has set up a ‘Strategy for smart, sustainable and inclusive growth’ (European Comission, 2010), where a crucial part of investment is to contribute to the education of the next generations: The aim is to reduce the share of early school leavers to below 10% and ensure that at least 40% of the younger generation obtain a tertiary degree.

@@

## 2. Dataset Description and Descriptive Statistics 
```{r a, include=FALSE}
########################## Read csv file########################
sm <- read.csv("/Users/deasysibarani/Desktop/Math_Statistics_Foundations/Group Project/student/student-mat.csv", sep=';') 
# academic performance is measured by the average of exam scores
sm[, "Score"] <- (sm[, "G1"] + sm[, "G2"] + sm[, "G3"])/3

############## Variable modification/derivation ################

# 1. father or mother job is teacher = 1
sm$f_or_mteach <- 0
sm[,"f_or_mteach"] <- as.character(sm[,"f_or_mteach"])
for (j in 1:nrow(sm)) {

  if (sm[j, "Fjob"] == "teacher") sm[j,"f_or_mteach"] <- 1
  if (sm[j, "Mjob"] == "teacher") sm[j,"f_or_mteach"] <- 1
}
sm[,"f_or_mteach"] <- as.numeric(sm[,"f_or_mteach"])

# 2. father or mother earns university degree = 1 
sm$f_or_medu <- 0
sm[,"f_or_medu"] <- as.character(sm[,"f_or_medu"])
for (j in 1:nrow(sm)) {

  if (sm[j, "Medu"] == 4) sm[j,"f_or_medu"] <- 1
  if (sm[j, "Fedu"] == 4) sm[j,"f_or_medu"] <- 1
}
sm[,"f_or_medu"] <- as.numeric(sm[,"f_or_medu"])

# 3. want to take higher education = 1

sm$higher_ed <- 0
sm[,"higher_ed"] <- as.character(sm[,"higher_ed"])
for (j in 1:nrow(sm)) {

  if (sm[j, "higher"] == "yes") sm[j,"higher_ed"] <- 1
  if (sm[j, "higher"] == "yes") sm[j,"higher_ed"] <- 1
}
sm[,"higher_ed"] <- as.numeric(sm[,"higher_ed"])

################################################################################

@@

#### d. Social Life and Health 
The social life section provides information on activities the student is involved in outside of school. This includes how much leisure time they have, whether they are in a relationship and how often they go out. Moreover, there is data on how much alcohol students consume and in what health condition they are in.

#### e. Grades
The grades are split into three grades: the first period (G1), second period (G2) and final grade (G3). For the purpose of this analysis, students academic performance will be measured as the average of all three grades. In addition, there is information on whether they have failed to pass any classes.

```{r e, echo=FALSE}
pie_failures
```

@@

As aforementioned, this paper will look into whether alcohol is one of the factors affecting academic performance. Our basic assumption is alcohol consumption has a negative impact on students grade (Ho). To test this hypothesis, multivariate regression analysis will be applied on the 395 sample of students described above. This will allow us to see the significant factors that contribute to academic performance, and check whether or not alcohol consumption is one of them. In this regression, grade is assigned as the dependent variable, and all other variables are treated as independent variable. If alcohol appeared as one of the significant variables in the regression, the null hypothesis can not be rejected, inferring that alcohol **is** one of the variables related to academic performance.

@@

## 4. Correlation Analysis and Regression

Prior to regression, there are several items that need to be inspected :

    a) Distribution of Response Variable

    If the distribution is highly skewed, linear regression might not give a good approximation or response varible.         Thus, variable transformation might be required, or applying other regression method might be more appropriate.

    b) Correlation Analysis

    Correlation between response and explanatory variables is required to get a general view of the variables   relationships. In addition, correlation between explanatory variables will also be assessed to avoid multicollinearity problem, as well as to provide other insights from the data.

### a. Distribution of Response Variable

```{r f, echo=FALSE}
sm[,"Score"] <- (sm[,"G1"] + sm[,"G2"] + sm[,"G3"])/3

ggplot(sm) + geom_density(aes(x=Score, fill = Score, colour = Score)) + labs(x = "Academic Score", y="% Frequency", title = "Academic Score of Secondary School Student")
```

As seen from the graph, the distribution is slightly left-skewed, but very close to normal distribution. Thus, multivariate regression method is appropriate to be applied to test the hypothesis, and no variable transformation required.

### b. Correlation Analysis

After performing correlation of all explanatory variables against grade (response variable), a few variables appear to have high correlation to grade :

1) Parent's Education
```{r g, echo=FALSE}
cor.test(sm[,"f_or_medu"],sm[,"Score"])
```
The correlation with grade is 0.23 With 95% confidence interval and p-value result of 3.586e-06, we can reject the null hypothesis of there is no correlation between these two variables. This means that parent's education has positive correlation with grade, and that students who have parents with university degree are more likely to have higher scores than those who do not.

2) Want to have higher education
```{r h, echo=FALSE}
cor.test(sm[,"higher_ed"],sm[,"Score"])
```
The correlation with grade is 0.189. With 95% confidence interval and p-value result of 0.0001517, we can reject the null hypothesis of there is no correlation between these two variables, and that grade does correlate positively with their aspiration to take higher education.

3) Number of past class failure
```{r i, echo=FALSE}
cor.test(sm$Score,sm$failures) 
```
The correlation with grade is -0.375. With 95% confidence interval and p-value result of 1.08e-14, we can reject the null hypothesis of there is no correlation between these two variables, and conclude that indeed, number of past class failure is negatively correlated with grade. students who have higher number of class failures in the past are more likely to have lower grade.

4) Frequency of going out with friends
```{r i1, echo=FALSE}
cor.test(sm$Score,sm$goout)
```

The correlation with grade is -0.154. With 95% confidence interval and p-value result of 0.002073, we can reject the null hypothesis of there is no correlation between these two variables, and conclude that indeed, the frequency of going out with friends are negatively correlated with grade. students who goes out more are more likely to have lower grade.

Alcohol does not have a significant correlation to grade, as shown in the correlation result result below :

```{r i2, echo=FALSE}
cor.test(sm$Score,sm$Walc)
```
The correlation shown is -0.08. With 95% confidence interval and p-value result of 8%, we can not reject the null hypothesis of there is no correlation between these two variables, and conclude that there is no correlation significant correlation between alcohol and students' academic performance.

suggestion: here we put remi's part. However, while performing correlation among explanatory variables .... (intro to remi's):

```{r i3, echo=FALSE}

###REMI's###

@@

### Regression

After assessing the correlations between all the variables, multivariate regression will be applied. Firstly, all variables are put in the regression. Then, variable with the highest p-value is eliminated from the model before the next regression is performed. This iteration is done until a model with satisfying R2 and consists of significant explanatory variables is derived.

The end result of the regression is shown in the figure below.

Initial Regression :

```{r i4, echo=TRUE}
#Initial Regression

reg <- lm(Score ~ school + sex + age + address + famsize + Pstatus + Medu + Fedu + 
                Mjob + Fjob + reason + guardian + traveltime + studytime + failures + 
                schoolsup + famsup + paid + activities + nursery + higher + internet + 
                romantic + famrel + freetime + goout + Dalc + Walc + health + absences +
                f_or_mteach + f_or_medu
                , data = sm)
summary(reg)
```

```{r j, include=FALSE}

@@

#To cross check on the final model and the significant variables, we use existing R library:
#library(MASS)
#step <- stepAIC(reg, direction="both")
```

**Figure 4. Final Regression Model**

```{r j1, echo=FALSE}
#Final Model

@@

From the regression result, we can see that alcohol consumption is not one of the significant variables that explains student academic performance. Instead, they are:

1) Sex

2) Address

3) Study time

4) Number of past class failure

5) Extra educational school support

6) Family educational support

7) Frequency of going out with friends

8) Parents level of education

The R2 is 21.74%, which means that the model can explain 21.74% of the variance in student's academic score. The residual plot shown below (figure ...) does not show a random dispersed around the x axis, which means the residual does not sum up to zero. There are other significant factors not captured in the dataset nor the model, that can contribute in explaining academic score. Some examples are IQ score, EQ score, nutrition, or time spent for physical exercise.

**Figure 5. Residual Plot**
